RILS-ROLS: Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-11-05 doc/paper.tex
Update: added Abstract

\author[1]{Aleksandar Kartelj}
\author[2]{Marko Djukanovi\'c}
	\address[1]{$kartelj@matf.bg.ac.rs$, \\  Faculty of Mathematics, University of Belgrade, Serbia}
    \address[2]{$ marko.djukanovic@pmf.unibl.org$,\\   Faculty of Natural Sciences and Mathematics, University of Banja Luka, Bosnia and Herzegovina}
	\begin{abstract}
		  \fxnote{TODO}
	\end{abstract}
	\maketitle

@@

\begin{itemize}
    \item The exact percentage rate of \textsc{Rils}--\textsc{Rols} algorithm  slightly decrease as the expected number of variables gets larger when no noise is presented in the data.
    \item The exact percentages drop off with the presence of a larger noise level. Thus, the higher the noise level, the smaller the exact percentages are. \fxnote{Marko: came up to here!}
    \item Interestingly, in the presence of noise, the highest exact percentages are not obtained on the problem instances whose exact models consists of a single variable, but two or three. We argue it by the fact that the early phase of the algorithm will more likely try to include additional variables in the solution at the cost of increasing the model accuracy to ensure improving the fitness value over iterations. In the presence of the zero noise, OLS method will more likely find a smaller model (on $\approx$ 50\% cases) with a high accuracy that fits to the (no-noisy) input data than the case of noisy data. When the data is noisy, the decision of adding more variables at the beginning is likely to be forced than just dealing with short-size solutions. By the algorithm's progression, it gets harder improving the best fitness value by adding variables, and other operations are more frequently involved than in previous iterations. %This conclusion also correlates to the shape of fitness function given by Equation~(\ref{eq:fitness}) where the first two product terms there are already small (close to 1).      \fxnote{TODO: need to be further expanded...}
	%Obtaining an accurate model with a few variables in the presence of a high level of nose in the data 
\end{itemize}

@@

\subsection{Statistical evaluation}
\fxnote{Marko: came up to here!}
In order to check the statistical significance of the obtained exact percentages and their differences between the studied approaches, we employed the statistical methodology comparing the results of  the 15 approaches on all problem instances from the both ground-truth benchmark sets, \textsc{Feynman} and \textsc{Strogatz}. Initially, Friedman’s test was separately executed for all competitor approaches.      In those cases in which the null hypothesis $H_0$ was rejected ($H_0$ states that there are no statistical differences between the results of the competitor approaches) pairwise comparisons are further performed by using the Nemenyi post-hoc test~\cite{pohlert2014pairwise}. The outcome is represented by means of critical difference (CD) plots. In a CD plot, the competitor approaches are placed on the horizontal axis according to their average ranking. Thereafter, the CD score is computed for a significance level of 0.05. If the difference is small enough, meaning that no statistical difference is detected, a horizontal bar linking statistically equal approaches is displayed.   The corresponding three (one for each level of noise) CD plots are shown in Figure~\ref{fig:cd-plots-exact-pcts}. The following conclusions may be drawn from there.

\begin{itemize}
	\item  Concerning instance problems where for the input data no noise is included, \textsc{Rils}-\textsc{Rols} method has the best average ranking in terms of   exact percentage rate.  The second best ranking is achieved by \textsc{AI-Feynman} algorithm. There is no statistically significant difference between these two approaches. However, these two approaches perform statistically significantly better than the third best (according the the average raking), that is with \textsc{Apf-Fe}. This conclusion also holds for the remaining approaches. 
	\item    Concerning instance problems where for the input data a (white) noise of 0.001 is included, \textsc{Rils}-\textsc{Rols} method has the best average ranking in terms of the exact percentage rate.   The second best ranking is achieved by \textsc{Afp-Fe} algorithm. However,  there is no statistically significant difference between these two approaches. The third best according approach is \textsc{Afp}, which perform statistically equal to \textsc{Afp-Fe}, but statistically worse that our \textsc{Rils}-\textsc{Rols} method. For all remaining methods, the later conclusion also holds.
	\item   Concerning instance problems where for the input data a (white) noise of 0.01 is included, again, \textsc{Rils-Rols} reaches the best average ranking; the second best approach is \textsc{gplearn}. The difference between the exact percentages delivered by these two approaches is significant in favor of \textsc{Rils-Rols} approach. This is also the case when \textsc{Rils-Rols} is compared to all the remaining approaches.

@@

\section{Conclusions and future work}\label{sec:conclusions}
  This paper deals with solving the well--known Symbolic regression (SR)  problem. This problem is a generalization of more interdisciplinary appearing  linear and logistic regression problems. SR problem have been intensively studied in the literature over the last 15 years, where a few dozens of algorithms were constructed, mainly based on Genetic programming and on the application of different kinds of neural networks. We proposed a meta-heuristic approach, called \textsc{Rils}-\textsc{Rols}. It is based on the Iterative local search (ILS) scheme. The first crucial concept within this scheme is utilizing the oridinary least square method to efficiently determine coefficients in the candidate solutions. Another key aspect is the use of a carefully constructed fitness function, which is a combination of three important scores: RMSE, $R^2$  scores and the size of model (solution).  Last but not least, the neighbor of solution in the local search is carefully defined that suits to small modifications of a three that represents solutions (models). \textsc{Rils-Rols} method is compared to the 14 other competitor algorithms from literature on the two ground truth benchmark sets from literature. Additionally, we introduced a non-biased benchmark set of randomly generated instances.  The experimental evaluation confirmed efficiently of out method which was able to produce the best average ranking in terms of exact percentage in comparison to all other competitors on all ground-truth instances, and in presence of three different levels of noise in the input data. In all cases, new best average-percentages are obtained in the literature by means of \textsc{Rils-Rols} method. In some cases, these differences are statistically significant in favor of our \textsc{Rils-Rols} method. Moreover, scalability and robustness of our method is  checked and confirmed on the randomly generated instances in terms of a few different criteria such as the size of model, the number of variables, average runtimes, etc.  \fxnote{TODO: case-study about large-scale data to be added}

  In the future work, one could think of  make a hybrid of \textsc{Rils-Rols} method with some other meta-heuristics, or replacing ILS scheme of \textsc{Rils-Rils} with, for example, Variable neighborhood search meta-heuristic. Additionally, as a proof of the concept, our tool will be provided to various research communities out of the are of Computer science, such as the ones from physics ot chemistry to seek for new hypotheses or o obtain insights on their experimental data.

@@

 \section{Overview of RILS-ROLS python package}\label{sec:appendix-1}

>>>>

\author[1]{Aleksandar Kartelj}
\author[2]{Marko Djukanovi\'c}
	\address[1]{$kartelj@matf.bg.ac.rs$, \\  Faculty of Mathematics, University of Belgrade, Serbia}
    \address[2]{$ marko.djukanovic@pmf.unibl.org$,\\   Faculty of Natural Sciences and Mathematics, University of Banja Luka, Bosnia and Herzegovina}
	\begin{abstract}
          This paper solves the well-known prominent problem of Symbolic regression. This problem have a wide range of applications.  For example, the obtained SR models could serve as insight towards  establishing  physical theory behind, thus helping establishing reasonable hypotheses.  Practical applications include revealing complex ecological dynamics in ecology, discovering mutations effects on protein stability in biology,  modeling analytic representations of the exciton binding energy in the energy science, and many applications in the material science. We propose a new meta-heuristic based approach, called \textsc{Rils}-\textsc{Rols} algorithm to solve the problem of Symbolic regression. There are a several concepts utilized within this algorithm: ($i$) Iterated local search scheme is used as the backbone; ($ii$) ordinary least square method is used within the local search procedure to make it efficient and to determine suitable coefficients in the models (solutions);  ($iii$) a novel fitness function is utilized which combines accuracy score, error score, and the size of model into one, more balanced value. The experiments are conducted on two ground-truth benchmark sets, \textsc{Feynman} and \textsc{Strogatz} comparing \textsc{Rils}-\textsc{Rols} to 14 other competitors from the literature under different levels of noise in the input data.  Our method obtains new best exact percentages on the both benchmark sets comparing to the literature approaches. Moreover, new unbiased randomly generated set of instances are introduced and scalability and robustness of our method is demonstrated on these instances in terms of various instance properties. 
	\end{abstract}
	\maketitle

@@

\begin{itemize}
    \item The exact percentage rate of \textsc{Rils}--\textsc{Rols} algorithm  slightly decrease as the expected number of variables gets larger when no noise is presented in the data.
    \item The exact percentages drop off with the presence of a larger noise level. Thus, the higher the noise level, the smaller the exact percentages are.
    \item Interestingly, in the presence of noise, the highest exact percentages are not obtained on the problem instances whose exact models consists of a single variable, but two or three. We argue it by the fact that the early phase of the algorithm will more likely try to include additional variables in the solution at the cost of increasing the model accuracy to ensure improving the fitness value over iterations. In the presence of the zero noise, OLS method will more likely find a smaller model (on $\approx$ 50\% cases) with a high accuracy that fits to the (no-noisy) input data than the case of noisy data. When the data is noisy, the decision of adding more variables at the beginning is likely to be forced than just dealing with short-size solutions. By the algorithm's progression, it gets harder improving the best fitness value by adding variables, and other operations are more frequently involved than in previous iterations. %This conclusion also correlates to the shape of fitness function given by Equation~(\ref{eq:fitness}) where the first two product terms there are already small (close to 1).      \fxnote{TODO: need to be further expanded...}
	%Obtaining an accurate model with a few variables in the presence of a high level of nose in the data 
\end{itemize}

@@

\subsection{Statistical evaluation}

In order to check the statistical significance of the obtained exact percentages and their differences between the competitor approaches, we employed the statistical methodology comparing the results of  the 15 approaches on all problem instances from the both ground-truth benchmark sets, \textsc{Feynman} and \textsc{Strogatz}. Note that in this case the exact percentage of each problem instance is given w.r.t.\  10 separately executed runs on that instance.

 Initially, Friedman’s test was separately executed for all competitor approaches.      In those cases in which the null hypothesis $H_0$ was rejected ($H_0$ states that there are no statistical differences between the obtained exact percentages of all competitors) pairwise comparisons are further performed by using the Nemenyi post-hoc test~\cite{pohlert2014pairwise}. The outcome is represented by means of critical difference (CD) plots. In a CD plot, the competitor approaches are placed on the horizontal axis according to their average ranking. Thereafter, the CD score is computed for a significance level of 0.05. If the difference is small enough, meaning that no statistical difference is detected, a horizontal bar linking statistically equal approaches is displayed.   The corresponding three (one for each level of noise) CD plots are shown in Figure~\ref{fig:cd-plots-exact-pcts}. The following conclusions may be drawn from there.

\begin{itemize}
	\item  Concerning the instance problems where no noise is included, \textsc{Rils}-\textsc{Rols} method has the best average ranking in terms of   exact percentage rate.  The second best ranking is achieved by \textsc{AI-Feynman} algorithm. There is no statistically significant difference between these two approaches. However, these two approaches perform statistically significantly better than the third best (according the the average raking), that is with \textsc{Apf-Fe}. This conclusion also holds for the remaining approaches, which are performing statistically significantly weaker than  \textsc{Rils}-\textsc{Rols} method. 
	\item    Concerning the instance problems with the included level of noise of 0.001, \textsc{Rils}-\textsc{Rols} method again obtains the best average ranking in terms of the exact percentage rate.   The second best ranking is achieved by \textsc{Afp-Fe} algorithm. However,  there is no statistically significant difference between these two approaches. The third best approach is \textsc{Afp}, which perform statistically equal to \textsc{Afp-Fe}, but statistically significantly worse that our \textsc{Rils}-\textsc{Rols} method. For all remaining methods, the later conclusion also holds.
	\item   Concerning the instance problems in the presence of the noise level of  0.01, again, \textsc{Rils-Rols} produces the best average ranking; the second best approach is \textsc{gplearn}. The difference between the exact percentages delivered by these two approaches is statistically significant in favor of \textsc{Rils-Rols} approach. This also holds when \textsc{Rils-Rols} is compared to all the remaining approaches.

@@

\section{Conclusions and future work}\label{sec:conclusions}

  This paper deals with solving the well--known Symbolic regression (SR)  problem. This problem is a generalization of more interdisciplinary appearing  linear and logistic regression problems. SR problem have been intensively studied in the literature over the last 15 years, where a few dozens of algorithms were constructed, mainly based on Genetic programming and on the application of different kinds of neural networks. We proposed a meta-heuristic approach, called \textsc{Rils}-\textsc{Rols}. Our approach is based on the Iterated local search (ILS) scheme. The first crucial concept within this scheme is utilizing the oridinary least square method to efficiently determine internal coefficients of the candidate solutions,  partially. Another key aspect is the use of a carefully constructed fitness function, which appears as  a combination of three important scores: RMSE, $R^2$  scores and the size of model (solution).  Last but not least, the neighbor of solution in the local search is carefully chosen that suits to small modifications of a three representing solutions (models). \textsc{Rils-Rols} method is compared to the 14 other competitor algorithms from literature on the two ground truth benchmark sets from literature, namely \textsc{Feynman} and \textsc{Strogatz}.  Additionally, we introduced a non-biased benchmark set of randomly generated instances, labelled by \textsc{Random}.  The experimental evaluation confirmed efficiently of our method which was able to produce the best average ranking results in terms of exact percentage wh comparison to all other competitors on the two ground-truth benchmark sets, in presence of the three different levels of noise in the input data. In all cases, new best exact percentages are obtained by means of \textsc{Rils-Rols} method. In some cases, these differences are statistically significant in favor of our \textsc{Rils-Rols} method. Moreover, scalability and robustness of this method is  checked and confirmed on unbiased \textsc{Random} benchmark set in terms of a few different criteria such as the size of model, the number of variables, average runtimes, etc.  \fxnote{TODO: case-study about large-scale data to be added}

  In the future work, one could think of constructing a hybrid of \textsc{Rils-Rols} method with some other meta-heuristics. Replacing ILS scheme of \textsc{Rils-Rils} with, a general scheme Variable neighborhood search meta-heuristic is a reasonable option to try out. Additionally, as a proof of the efficiency of our algorithm, our tool should be provided to various research communities out of the are of Computer science, such as the ones from physics or chemistry to help for  seeking new hypotheses or to obtain valuable insights on their experimental evaluations.

@@

 \section{Overview of RILS-ROLS python package}\label{sec:appendix-1}
  \fxnote{TODO}
